Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-01 Delaneyetal2024.tex
overleaf grammer

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones, however.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \citep{rothlisberger1972}.
As a result  the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow over short periods (hours, days).
Changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \citep[Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; ][]{alley1997}.

@@

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of theses studies discuss the disparity between water discharge and sediment export, highlighting the non-linear relationship between the two, along with the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited, or hydraulically controlled, regime.
The response of sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in these glacierized catchments systems \citep[e.g.][]{ganti2016,mancini2023}.
It is important that this relationship is clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

@@

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our choice to evaluate shear stress omits selecting a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.

@@

The shear stress, $\tau_g$, between the water and the channel bed results from the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
% 
where $v_g = \frac{Q}{S_g}$ is the water velocity.
% 
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
% 
This value is used to establish the integrated shear stress across the channel $w_g\tau_g$.

@@

The hydraulics parameterization presented in \citet{tucker1997} is implemented to represent the subaerial channel.
This model uses mass conservation and the Darcy-Weisbach relationship and assumes that the channel is sufficiently wide compared to its depth.
This latter assumption means that the hydraulic radius is well approximated by the flow depth.
Therefore, the resulting shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_f^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_f$ is the friction factor for subaerial channels \citep{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
% 
where $k$ is a constant and $\alpha=\frac{1}{3}$ is a commonly chosen exponent \citep{leopold1953}.
Also following the Darcy-Weisbach, subaerial water velocity, $v_f$, is given as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_f\,\rho_w}}.
\end{equation}
% 
As above, the width-integrated shear stress is $w_f\tau_f$.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line is given refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine differing water discharge variability by smoothing the water discharge over different period from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal.
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

The effects of channel shape, gradient, and roughness is established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$, and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$ \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
To be accepted, subglacial model runs must experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in the channel area.

@@

In the second experiment, the models are applied to a reference test case for each glacier.
These experiments assumes  a subglacial channel with $\beta=\frac{\pi}{6}$ and a subaerial channel with $\alpha = \frac{1}{3}$ and slope of $0.02$ (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \citep[$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$][]{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
% 
To test covariance between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation.
This metric accounts for the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear relationship between sediment transport capacity and hydrology.

@@

Table~\ref{tab:Qs} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ \& $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than th other transport relations.

The sediment discharge in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.
However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely  much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in water discharge.
The timescale of  diurnal melt or severe rain or melt  are too short  for the channel to reach a steady state.
In these cases with high water discharge variability, channels operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that  sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}.

@@

Smaller values of $\alpha$, or a channel cross section closer to a slot-canyon, result in greater variability in velocity and shear stress (Figure~\ref{fig:params}).
In both cases, the variability of small values of $\alpha$ does not approach that of the subglacial channels.
Yet, larger values of $\alpha$ can result in greater variability in width-integrated shear stress, as greater channel width changes increase the width-integrated shear stress.
Additionally, steeper subaerial channel slopes result in greater variability in both shear stress and width-integrated shear stress but only approach the  subglacial channel's variability in the \alpine case (Figure~\ref{fig:params}). 
We note that these parameter values span a commonly accepted range.
Therefore , we do not anticipate a scenario where variability in the subaerial system would exceed the subglacial system with this two hydrologic forcings.

@@

The second numerical experiment aims to quantify the sources of increased variability in the subglacial model outputs as they exhibit different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because there subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig5.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
  }
  \label{fig:model_outs}
\end{figure}

@@

Several aspects of the models and experiment design limit make comparing the model outputs to field observations difficult and make comparison of the two systems difficult.
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, sediment access is not considered here.
In reality, processes such as sediment access are highly important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here making comparison more direct by isolating the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important, spatially distributed processes.

Comparing model outputs between the subaerial and subglacial case is difficult due to the different gradients in the two systems, with the subglacial one being much steeper \citep{alley1997}.
The water velocities are similar in both channel types (Figure~\ref{fig:model_outs}).
However, the shear stresses and width-integrated shear stresses are much smaller in subaerial channels.
We note that in the ensemble runs in Section~\ref{sect:ensemble}, the parameters' range tested covers a likely range of viable shear stresses in both subglacial and subaerial channels.
The greater sediment transport capacity in the subglacial channels implies that sediment grain-size underneath subglacial channels is likely larger that subaerial counterparts.

@@

A sporadic relationship between subglacial sediment transport capacities and water discharge occurs due to hysterisis in subglacial channel size (Figure~\ref{fig:Qw_vari}).
This characteristic limits the use of water discharge as an indicator of sediment discharge capacity in these systems.
As a result, characteristics such as bankfull or effective water discharge that link geomorphic work to hydro-climatic conditions could have limited meaning in evaluating subglacial sediment transport \citep{wolman1960,lenzi2006}.
A glacier's sediment transport capacity is impacted by the ice thickness controlling the channel closure rate and the glacier's surface slope, in addition to water discharge and sediment size \citep[Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,gimbert2016,stevens2022,walder1994}.
This multitude of processes lies in contrast to many subaerial channels, where transport capacity typically responds to water discharge, sediment size, channel shape,  and hydraulic gradient.
The latter of these parameters can remain relatively stable over the years or longer \citep[Section~\ref{sect:fluv}; e.g.][]{tucker1997}.
Strong correlations between water discharge and sediment export in glacier systems could indicate  other processes, such as increased sediment access \citep{zhang2022}.
Furthermore, the observed correlation between water discharge and sediment discharge far downstream likely represents subaerial transport processes, especially with respect to bedload \citep{mancini2023}.

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson river's water discharge into the Kangerlussuaq fjord \citep{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).

Water discharge measurements at the less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decrease model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with subglacial channel, when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Section ~\ref{sect:scaling}).
Yet, such strong relationship would not represent the impact of shorter term fluctuations in water discharge.
Note, that if basal melt largely dominates, then water discharge would evolve slowly with respect to the subglacial channel size.

Clockwise hysterisis loops between sediment concentration and water discharge suggest that sediment availability is reduced over the event scale \citep[][]{williams1989}.
These loops are often observed in glacierized catchments and used to suggest that access to subglacial sediment has been limited \citep[e.g.][]{collins1979,willis1996,richards2003,stott2007,delaney2018}.
Results here suggest that clockwise hysterisis could also be expected in transport-limited regimes in many subglacial environments.
Here, the smaller channel size on the rising limb would result in greater sediment transport capacity and sediment uptakes.
Larger channel size would reduce sediment transport capacity for an equivalent water discharge on the falling limb, creating clockwise hysterisis.

@@

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, in most systems an incoherent relationship between water and subglacial sediment discharge could be expected.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Due to evolving subglacial channel size, the timing of peak sediment discharge capacity in subglacial channels occurs before peak water discharge during an event, whereas it is coincident in subaerial ones.
Our analysis shows, however, that covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to often not be the case, except where basal melt dominates.

>>>>

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones, however.
The size of subglacial channels is controlled by the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \citep{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow over short periods (hours, days).
Changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \citep[Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; ][]{alley1997}.

@@

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, highlighting the non-linear relationship between the two, along with the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited, or hydraulically controlled, regime.
The response of sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in these glacierized catchment systems \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

@@

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our choice to evaluate shear stress omits the selection of a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.

@@

The shear stress, $\tau_g$, between the water and the channel bed results from the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
% 
where $v_g = \frac{Q}{S_g}$ is the water velocity.
% 
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
% 
This value establishes the integrated shear stress across the channel $w_g\tau_g$.

@@

The hydraulics parameterization presented in \citet{tucker1997} is implemented to represent the subaerial channel.
This model uses mass conservation and the Darcy-Weisbach relationship and assumes that the channel is sufficiently wide compared to its depth.
This latter assumption means that the hydraulic radius is well approximated by the flow depth.
Therefore, the resulting shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_f^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_f$ is the friction factor for subaerial channels \citep{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
% 
where $k$ is a constant and $\alpha=\frac{1}{3}$ is a commonly chosen exponent \citep{leopold1953}.
Also following the Darcy-Weisbach, subaerial water velocity, $v_f$, is given as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_f\,\rho_w}}.
\end{equation}
% 
As mentioned above, the width-integrated shear stress is $w_f\tau_f$.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine differing water discharge variability by smoothing the water discharge over different periods from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal.
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

The effects of channel shape, gradient, and roughness are established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$, and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$ \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
To be accepted, subglacial model runs must experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in the channel area.

@@

In the second experiment, the models are applied to a reference test case for each glacier.
These experiments assumes  a subglacial channel with $\beta=\frac{\pi}{6}$ and a subaerial channel with $\alpha = \frac{1}{3}$ and slope of $0.02$ (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \citep[$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$][]{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
% 
To test the covariance between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation.
This metric accounts for the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear relationship between sediment transport capacity and hydrology.

@@

Table~\ref{tab:Qs} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ \& $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

The sediment discharge in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely much higher in a steady-state R-channel.

Furthermore, R-channels will rarely operate in a steady state as variations in water discharge.
The timescale of diurnal melt or severe rain or melt is too short for the channel to reach a steady state.
In these cases with high water discharge variability, channels operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}.

@@

Smaller values of $\alpha$, or a channel cross-section closer to a slot canyon, result in greater variability in velocity and shear stress (Figure~\ref{fig:params}).
In both cases, the variability of small values of $\alpha$ does not approach that of the subglacial channels.
Yet, larger values of $\alpha$ can result in greater variability in width-integrated shear stress, as greater channel width changes increase the width-integrated shear stress.
Additionally, steeper subaerial channel slopes result in greater variability in both shear stress and width-integrated shear stress but only approach the subglacial channel's variability in the \alpine case (Figure~\ref{fig:params}). 
We note that these parameter values span a commonly accepted range.
Therefore, we do not anticipate a scenario where variability in the subaerial system would exceed the subglacial system with these two hydrologic forcings.

@@

The second numerical experiment aims to quantify the sources of increased variability in the subglacial model outputs as they exhibit different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig5.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
  }
  \label{fig:model_outs}
\end{figure}

@@

Several aspects of the models and experiment design limit make comparing the model outputs to field observations difficult and make comparison of the two systems difficult.
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, sediment access is not considered here.
In reality, processes such as sediment access are highly important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here make the comparison more direct by isolating the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important, spatially distributed processes.

Comparing model outputs between the subaerial and subglacial case is difficult due to the different gradients in the two systems, with the subglacial one being much steeper \citep{alley1997}.
The water velocities are similar in both channel types (Figure~\ref{fig:model_outs}).
However, the shear stresses and width-integrated shear stresses are much smaller in subaerial channels.
We note that in the ensemble runs in Section~\ref{sect:ensemble}, the parameters' range tested covers a likely range of viable shear stresses in both subglacial and subaerial channels.
The greater sediment transport capacity in the subglacial channels implies that sediment grain size underneath subglacial channels is likely larger than subaerial counterparts.

@@

A sporadic relationship between subglacial sediment transport capacities and water discharge occurs due to hysteresis in subglacial channel size (Figure~\ref{fig:Qw_vari}).
This characteristic limits the use of water discharge as an indicator of sediment discharge capacity in these systems.
As a result, characteristics such as bankfull or effective water discharge that link geomorphic work to hydro-climatic conditions could have limited meaning in evaluating subglacial sediment transport \citep{wolman1960,lenzi2006}.
A glacier's sediment transport capacity is impacted by the ice thickness controlling the channel closure rate and the glacier's surface slope, in addition to water discharge and sediment size \citep[Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,gimbert2016,stevens2022,walder1994}.
This multitude of processes lies in contrast to many subaerial channels, where transport capacity typically responds to water discharge, sediment size, channel shape,  and hydraulic gradient.
The latter of these parameters can remain relatively stable over the years or longer \citep[Section~\ref{sect:fluv}; e.g.][]{tucker1997}.
Strong correlations between water discharge and sediment export in glacier systems could indicate other processes, such as increased sediment access \citep{zhang2022}.
Furthermore, the observed correlation between water discharge and sediment discharge far downstream likely represents subaerial transport processes, especially with respect to bedload \citep{mancini2023}.

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson River's water discharge into the Kangerlussuaq fjord \citep{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).

Water discharge measurements at less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decreased model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with the subglacial channel when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Section ~\ref{sect:scaling}).
Yet, such a strong relationship would not represent the impact of shorter-term fluctuations in water discharge.
Note, that if basal melt largely dominates, then water discharge would evolve slowly with respect to the subglacial channel size.

Clockwise hysteresis loops between sediment concentration and water discharge suggest that sediment availability is reduced over the event scale \citep[][]{williams1989}.
These loops are often observed in glacierized catchments and used to suggest that access to subglacial sediment has been limited \citep[e.g.][]{collins1979,willis1996,richards2003,stott2007,delaney2018}.
Results here suggest that clockwise hysteresis could also be expected in transport-limited regimes in many subglacial environments.
Here, the smaller channel size on the rising limb would result in greater sediment transport capacity and sediment uptakes.
A larger channel size would reduce sediment transport capacity for an equivalent water discharge on the falling limb, creating clockwise hysteresis.

@@

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, in most systems, an incoherent relationship between water and subglacial sediment discharge could be expected.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Due to evolving subglacial channel size, the timing of peak sediment discharge capacity in subglacial channels occurs before peak water discharge during an event, whereas it is coincident in subaerial ones.
Our analysis shows, however, that covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to often not be the case, except where basal melt dominates.
